The effects of alignment error and alignment filtering on the \sw detection of positive selection

### 2012-02-06 projects/slrsim/paper/739_main_nographics.tex

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
\usepackage[a4paper,hmargin=2.5cm,vmargin=2.5cm]{geometry}
\usepackage{color}

@@

\newcommand{\mtwo}{PAML M2a\xspace}
\newcommand{\meight}{PAML M8\xspace}
\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\tprf}{TPR$_{5\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\emph{Drosophila}}\xspace}

\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}
%%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in vertebrates}
\author{{Gregory Jordan and Nick Goldman}}
%\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
\maketitle
%\authorlist{Jordan and Goldman}
%\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}

%\mbeabstract{% 
\section*{Abstract}
\textcolor{red}{
Alignment filters are commonly applied to try to reduce errors
resulting from misalignment, but the prevalence of such errors and the
ability of filters to reduce them is not well understood.
% 
Focusing on the sitewise detection of positive selection across a wide
range of divergence levels and indel rates, we performed simulation
experiments to quantify the false positives and false negatives
introduced by alignment error and the ability of alignment filters to
improve performance.
%
We found that some aligners led to many false positives while others
resulted in very few.  False negatives were a problem for all
aligners, increasing with divergence. Of the aligners tested, PRANK's
codon-based alignments consistently performed the best and ClustalW
performed the worst.
%
Of the filters tested, GUIDANCE performed the best and Gblocks
performed the worst. Although some filters showed good ability to
reduce the error rates from ClustalW and MAFFT alignments, none were
found to substantially improve the performance of PRANK alignments
under most \textcolor{red}{conditions}.
% 
Our results revealed distinct trends in error rates and power levels
for aligners and filters within a biologically plausible parameter
space, and controls using the true alignment and an optimal filtering
method suggested that performance improvements could be gained by
improving aligners or filters to reduce the prevalence of false
negatives, especially at higher divergence levels and indel rates.
}
%\keywords{Alignment, positive selection, alignment filtering, indels}

%\email{greg@ebi.ac.uk}

%\mbestyle

@@

A major concern in the detection of protein positive selection is that
the effect of alignment error is not well characterized. Intuitively,
one might expect alignment error to result mainly in an increased
number of false positives, as the spurious alignment of nonhomologous
codons on average would result in a high number of apparent
non-synonymous substitutions and a low number of synonymous
substitutions (since a randomly chosen pair of codons are more likely
to be nonsynonymous than synonymous). However, false negatives may
also be introduced, either through over-alignment of synonymous codons
(increasing the synonymous substitution rate) or non-alignment of
homologous codons (causing reduced power due to less evolutionary
information). Since different aligners employ a variety of algorithms,
evolutionary models, and heuristic optimizations
\citep{Notredame2007Recent}, each program may be more or less prone to
different types of alignment error, causing potentially large
variations in the nature and magnitude of its impact on the detection
of positive selection.

\input{table1.txt}

The protein structure and the evolutionary divergence of a dataset may
also contribute to the effects of alignment error. Differently
structured protein regions show variable tolerance to biological
indels, with indels more common in extracellular and transmembrane
proteins than in highly folded enzymes and housekeeping genes
\citep{delaChaux2007DNA}. This suggests that well-folded protein
regions will experience fewer biological indels---and will therefore
be less susceptible to alignment error---than less structured
regions.

@@

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks program applies a set of
heuristic criteria to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection}
while a number of aligners such as T-Coffee
\citep{Notredame2000TCoffee} and PRANK
\citep{Loytynoja2008PhylogenyAware} produce estimates of alignment
confidence or reliability. GUIDANCE, which measures the robustness of
alignment regions to perturbations in the guide tree used for
progressive alignment, has also been proposed as an alignment
confidence score \citep{Penn2010Alignment}. Unfortunately, despite
their widespread use, the impact of the many available alignment
scoring and filtering methods on phylogenetic and evolutionary
analyses has not been well studied. Even for a single filtering
program, Gblocks, results have been contradictory: one
simulation-based study found that it improved the phylogenetic signal
\citep{Talavera2007Improvement} while an empirical study across a wide
range of taxa found that Gblocks-filtered alignments produced worse
phylogenetic trees than unfiltered alignments
\citep{Dessimoz2010Phylogenetic}. \textcolor{red}{A recent study suggested that the
benefit of alignment filtering (in terms of improved accuracy)
outweighs the cost (in terms of reduced power) when applied to
detecting positive selection, but the analysis was limited to a small
range of possible evolutionary scenarios, as discussed below
\citep{Privman2011Improving}.}  With the application of published
filtering methods to alignments before testing for positive selection
becoming standard practice
\citep{Studer2008Pervasive,Aguileta2009Rapidly}, \textcolor{red}{continued investigation into the}
potential benefits of alignment filtering to the detection of positive selection
would seem well-warranted.

This paper aims to use a simulation framework to incorporate alignment
error and alignment filtering into estimates of the error rate and
power of \sw evolutionary inference of positive selection. Our
approach \textcolor{red}{is modeled on that of} Anisimova et al.
\citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
alignments to evaluate methods for detecting \sw positive selection,
considering each site to be an independent hypothesis test producing a
positive or negative result. \textcolor{red}{As we simulate codon
  alignments with indels and calculate inferred alignments using a
  small but diverse sample of aligners and alignment filters, the
  experimental design also shares some features with Fletcher and
  Yang's \citeyearpar{Fletcher2010Effect} investigation of the
  performance of the branch-site method for detecting positive
  selection under alignment error and Privman et al.'s
  study of the effect of alignment
  filtering on detecting positive selection in HIV-1 genes. However,
  the present work differs from previous studies in a number of
  important ways.}

\textcolor{red}{Unlike Fletcher and Yang, here we focus on \sw
  detection of sitewise positive selection occurring throughout a
  phylogeny (as opposed to genewise detection of selection acting at
  specific branches) and we evaluate the impact of a number of
  alignment filtering methods on the \sw analysis. Thus, the
  biological hypothesis being investigated is different, and the
  addition of filtering methods adds another layer of complexity to
  our analysis.}

\textcolor{red}{Privman et al., whose simulations centered on the
  evolutionary chracteristics of three HIV-1 genes, concluded that
  alignment filtering improves the performance of positive selection
  inference by removing false positives. While this may be true for
  the three genes (and perhaps for HIV-1 in general), HIV-1 is known
  to evolve with widespread positive selection in the human host
  \citep{Yang2003Widespread}. Results valid for these three HIV-1
  genes may not be widely applicable to large-scale vertebrate and
  mammalian comparative datasets, which exhibit less adaptive
  evolution \citep{Kosiol2008Patterns} and which comprise a larger
  diversity of protein structures and a wider range of species
  divergence levels.}

\textcolor{red}{We hypothesize that the divergence level and indel
  rate, two highly variable and important evolutionary factors, may
  strongly affect the performance of methods for alignment and
  detection of selection. Accordingly, our simulations encompass a
  wide range of biologically plausible indel rates and divergence
  levels while fixing other parameters at values typical to those
  encountered in the \sw analysis of vertebrate gene
  families. Although our main purpose is to understand the behaviour
  of various aligners and filters within our range of simulations, we
  will make use of the above-noted methodological similarities in
  order to highlight aspects of our results which agree with or
  contradict the literature where possible.}

@@

\input{table2.txt}

@@

An overview of the simulation parameters used in this study can be
found in Table \ref{table_1}. Three rooted trees were used to guide
the simulation of protein-coding DNA alignments: the artificial
6-taxon tree used by Anisimova et al.
\citeyearpar{Anisimova2001Accuracy} and Massingham and Goldman
\citeyearpar{Massingham2005Detecting} rooted at its midpoint, the
17-taxon vertebrate $\beta$-globin tree from Yang et
al. \citeyearpar{Yang2000CodonSubstitution} and the 44-taxon
vertebrate tree used by the ENCODE project
\citep{2007Identification,Nikolaev2007Early}. Trees, shown with their
original branch lengths in Figure \ref{fig_1}, were scaled to
comparable divergence levels by normalizing their mean path length
(MPL), defined as the root-to-tip branch length averaged across all
lineages in the tree. We simulated alignments with MPL divergence
between 0.05 and 2.0 synonymous substitutions per synonymous site,
spanning the range of evolutionary divergences observed in several
clades of organisms with fully-sequenced genomes (Table
\ref{table_2}).
\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.5]{fig1.pdf}
\includegraphics[scale=0.8]{fig1.pdf}
\end{center}
\caption{Phylogenetic trees used for simulation and analysis. The
  original scale for each tree is indicated by a scale bar, but trees
  were scaled to equal mean path length (MPL) divergence levels for
  simulation. (a) A 6-taxon artificial tree used in previous
  simulations
  \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (b) A tree
  estimated from $\beta$-globin genes of 17 vertebrates and used in
  previous empirical analyses and simulation studies
  \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (c) The
  44-species tree used by the ENCODE project
  \citep{2007Identification,Nikolaev2007Early}. The nodes indicated by
  arrows were used as the reference species when comparing the true
  and inferred alignment (see Methods).}
\label{fig_1}
\end{figure}

@@

The distribution of \sw selective pressures (embodied by the parameter
$\omg$, the ratio of the rate of non-synonymous substitution to the
rate of synonymous substitution) was modeled with a log-normal
distribution derived from a maximum-likelihood fit to a large dataset
of \sw selective pressures estimated from mammalian gene trees
(Lindblad-Toh et al. \citeyear{LindbladToh2011High}; log-normal parameters shown in Table
\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
sites having $\omg>1$, is consistent with the structure-based
expectation of many protein sites under purifying selection and few
under neutral selection or positive selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. INDELible's general
discrete model of \sw $\omg$ variation was used to approximate the
log-normal distribution by splitting the probability density into 50
equally-spaced bins between $\omg$ values of 0 and 3, with the highest
bin containing the probability density for all values $\omg>3$.

Branch lengths for each of the simulation trees were scaled before
simulation to correct for the difference between our definition of
branch lengths as the number of synonymous substitutions per
synonymous site ($dS$) and INDELible's interpretation of branch length
as the average number of substitutions per codon ($t$)
\citep{Fletcher2010Effect}. They are related approximately by
$t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and $S$ are the
proportion of nonsynonymous and synonymous sites and $\bar{\omg}$ is
the mean $\omg$ across all sites. $S$ is approximately 0.3 when
$\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio
for our chosen distribution is 0.277, yielding a $dS$-to-$t$
conversion factor of 1.48 for all simulations performed.

@@

Alignments were inferred using four alignment algorithms chosen for
their widespread use or demonstrated accuracy: ClustalW v1.82
\citep{Thompson1994ClustalW}, MAFFT \citep{Katoh2005MAFFT} and two
variants of PRANK \citep{Loytynoja2005From} based on an amino acid
model (subsequently referred to as \pranka{}) or an empirical codon
model (subsequently referred to as \prankc{}). Unaligned amino acid
sequences were given as input to all alignment programs (except
\prankc{}, which was provided the unaligned DNA sequences) and all
software was run using default parameters with the true phylogenetic
tree given as input where possible.

@@

GUIDANCE and T-Coffee produce individual scores for each residue,
allowing individual residues to be masked instead of entire
columns. \textcolor{red}{Privman et al. \citeyearpar{Privman2011Improving} found a
residue-based filter to be more effective than its column-based
equivalent, so we filtered residues instead of alignment columns where
possible}. GUIDANCE generates many replicate alignments, each using a
slightly perturbed guide tree, with either MAFFT or \pranka as the
bootstrap aligner. The program then assigns to each residue from the
input alignment a score from 0 to 1 based on how consistently it was
placed in the replicate alignments. In order to maximize the
similarity between the input aligner and the bootstrap aligner, we ran
GUIDANCE with 100 MAFFT replicates when filtering ClustalW alignments
and with 30 \pranka replicates when filtering \prankc
alignments. T-Coffee calculates the residue-wise consistency between
an input multiple alignment and independently calculated pairwise
alignments \citep{Notredame2003Using}, rounding and normalizing
residue scores into integers between 0 and 9. T-Coffee was run using
its default settings and the {\em evaluate\_mode
  --output=score\_ascii} command-line parameters to output alignment
scores.

@@

\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.75]{fig2.pdf}
\includegraphics[scale=1.25]{fig2.pdf}
\end{center}
\caption{Alignments were simulated without indels for three tree
  shapes and analyzed with SLR, \meight, or \mtwo. Fifty replicate
  alignments were simulated for each data point. The performance of
  each analysis method, as measured by four summary statistics, is
  plotted as a function of the mean path length (MPL) divergence. From
  top to bottom: true positive rate (TPR) at the recommended cutoff
  threshold (0.95 for PAML and 3.84 for SLR); false positive rate
  (FPR) at the recommended cutoff threshold; TPR at a 1\% FPR
  threshold; Pearson's correlation coefficient between the true and
  inferred sitewise $\omg$.}
\label{fig_2}
\end{figure}

\section*{Results and Discussion}
\subsection*{The Performance of Three Methods for Detecting \Sw Positive Selection}
We first evaluated the ability of three \sw methods, \mtwo,
\meight and SLR, to accurately estimate \sw $\omg$ values and to
detect positive selection under a range of tree lengths in the absence
of alignment error. Figure \ref{fig_2} shows the TPR, FPR, \tpr{} and \sw $\omg$
correlation over a range of MPLs for each of the three simulation
trees.

The detection power and $\omg$ correlation were weakest at low
divergence levels for all methods and all trees due to the low amount
of evolutionary information, as observed in previous simulations
\citep{Anisimova2002Accuracy}. We found a positive correlation between
tree size and detection power, with the highest performance in the
44-taxon tree. Power generally increased monotonically with
divergence, except for the 6-taxon tree which saw its maximum
performance at moderate divergence levels (MPL 0.5--1.0) and began
decreasing at higher values. The downward trend in the 6-taxon tree
was likely due to the impact of saturation of synonymous sites in the
very long branches present in such a sparse tree at high divergence
levels. With lower average branch lengths at equivalent MPLs, the two
larger trees showed no signs of decreased performance even at a MPL of
2 substitutions per site, which is greater than any of the divergence
levels found in groups of commonly analyzed vertebrate, insect and
fungal species (Table \ref{table_2}).

@@

In this paper, we investigated the performance of sitewise detection
of positive selection under a range of tree sizes, indel rates, and
divergence levels, using simulation parameters designed to approximate
the analysis of typical mammalian protein-coding genes. We
evaluated the ability of four alignment methods and three alignment
filtering methods to produce alignments for detecting positively
selected sites, using the FPR, TPR, and error-controlled \tpr of the
sitewise detection of positive selection as our main measures of
performance.

>>>>

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
\usepackage[a4paper,hmargin=2.5cm,vmargin=2.5cm]{geometry}
\usepackage{color}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{authblk}

@@

\newcommand{\mtwo}{PAML M2a\xspace}
\newcommand{\meight}{PAML M8\xspace}
\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\tprf}{TPR$_{5\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\emph{Drosophila}}\xspace}
%\newcommand{\tcr}[1]{\textcolor{red}{#1}}
%\newcommand{\tcrr}[1]{\textcolor{red}{#1}}
\newcommand{\tcr}[1]{#1}
\newcommand{\tcrr}[1]{#1}
\newcommand{\nsyn}{nonsynonymous\xspace}
\newcommand{\syn}{synonymous\xspace}
\newcommand{\nh}{non-homologous\xspace}
%\newcommand{\hidegraphics}[2]{\includegraphics[#1]{#2}}
\newcommand{\hidegraphics}[2]{}

\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}
\date{October 26, 2011}
\author[1]{Gregory Jordan}
\author[1,*]{Nick Goldman}
\affil[1]{EMBL-EBI, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}
\affil[*]{To whom correspondence should be addressed: Nick Goldman, Tel:~+44~(0)1223~492~530; Fax:~+44~(0)1223~494~468; E-mail:~goldman@ebi.ac.uk}

\maketitle

\subsection*{Keywords}
Alignment error, positive selection, codon models, alignment filtering

\subsection*{Running head}
Alignment error and detecting sitewise selection

\subsection*{Abbreviations used}
MPL---mean path length; TPR---true positive rate; FPR---false positive rate

\newpage

\section*{Abstract}
When detecting positive selection in proteins, the prevalence of
errors resulting from misalignment and the ability of alignment
filters to mitigate such errors are not well understood, but filters
are commonly applied to try to avoid false positive results.
% 
Focusing on the sitewise detection of positive selection across a wide
range of divergence levels and indel rates, we performed simulation
experiments to quantify the false positives and false negatives
introduced by alignment error and the ability of alignment filters to
improve performance.
%
We found that some aligners led to many false positives while others
resulted in very few.  False negatives were a problem for all
aligners, increasing with sequence divergence. Of the aligners tested, PRANK's
codon-based alignments consistently performed the best and ClustalW
performed the worst.
%
Of the filters tested, GUIDANCE performed the best and Gblocks
performed the worst. Although some filters showed good ability to
reduce the error rates from ClustalW and MAFFT alignments, none were
found to substantially improve the performance of PRANK alignments
under most conditions.
%
Our results revealed distinct trends in error rates and power levels
for aligners and filters within a biologically plausible parameter
space. With the best aligner, a low false positive rate was maintained
even with extremely divergent, indel-prone sequences. Controls using
the true alignment and an optimal filtering method suggested that
performance improvements could be gained by improving aligners or
filters to reduce the prevalence of false negatives, especially at
higher divergence levels and indel rates.

%\keywords{Alignment, positive selection, alignment filtering, indels}

@@

A major concern in the detection of protein positive selection is that
the effect of alignment error is not well characterized. Intuitively,
one might expect alignment error to result mainly in an increased
number of false positives, as the spurious alignment of \nh codons on
average would result in a high number of apparent \nsyn substitutions
and a low number of \syn substitutions (since two randomly chosen
codons are more likely to be \nsyn than \syn).  \tcr{However, false
  negatives may also be introduced, either through the introduction of
  \syn but \nh codons into a positively-selected site (thus reducing
  power due to an inflated \syn substitution rate) or through the
  failure to align truly homologous codons at a positively-selected
  site (reducing power due to less evidence for positive selection at
  that site).}  Since different aligners employ a variety of
algorithms, evolutionary models, and heuristic optimizations
\citep{Notredame2007Recent}, each program may be more or less prone to
different types of alignment error, causing potentially large
variations in the nature and magnitude of its impact on the detection
of positive selection. \tcr{Different aligners may also be designed
  for different downstream applications, such as phylogenetic
  inference or functional annotation \citep{Morrison2009}, making the
  optimal choice of aligner potentially dependent on the way in which
  the resulting alignment will be used. Here we focus on the sitewise
  detection of positive selection.}

\begin{table}
\caption{Parameter Values Used in Simulations}
\begin{threeparttable}
\footnotesize
\begin{tabular}{lllllllll}
\toprule
 \multicolumn{3}{c}{Tree} & \multicolumn{3}{c}{Insertions and Deletions} & \multicolumn{3}{c}{$\omega$ Distribution} \\
\cmidrule(r){1-3} \cmidrule(r){4-6} \cmidrule(r){7-9}
 & & & Size & Mean Length & & & & \\
Taxa & Source & MPL & Distribution & (Std. Dev.) & Rate & Shape & Mean & $p (\omega>1)$ \\
\midrule
6 & Artificial & & power law & & & lognormal & & \\
17 & $\beta$-globin & 0.05--2.0 & decay: 1.8 & 3.33 (5.51) & 0--0.2 & log mean: -1.864 & 0.277 & 0.06 \\
44 & Vertebrates & & max length: 40 & & &  log SD: 1.201 & & \\
\bottomrule
\label{table_1}
\end{tabular}
\begin{tablenotes}
\item [] N{\sc ote}.---{\hskip -.1pt}MPL is the mean path length of the tree in units of substitutions per synonymous site. Indel lengths are measured in units of codons, and the indel rate is defined as the number of insertion \& deletion events per substitution.
\end{tablenotes}
\end{threeparttable}
\end{table}

\tcr{In addition, we expect the protein structure and the
  evolutionary divergence of a dataset to contribute to the effects of
  alignment error}. Differently structured protein regions show
  variable tolerance to biological indels, with indels more common in
  extracellular and transmembrane proteins than in highly folded
  enzymes and housekeeping genes \citep{delaChaux2007DNA}. This
  suggests that well-folded protein regions will experience fewer
  biological indels---and will therefore be less susceptible to
  alignment error---than less structured regions.

@@

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks program applies a set of
heuristic criteria to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection}
while a number of aligners such as T-Coffee
\citep{Notredame2000TCoffee} and PRANK
\citep{Loytynoja2008PhylogenyAware} produce estimates of alignment
confidence or reliability. GUIDANCE, which measures the robustness of
alignment regions to perturbations in the guide tree used for
progressive alignment, has also been proposed as an alignment
confidence score \citep{Penn2010Alignment}. Unfortunately, despite
their widespread use, the impact of the many available alignment
scoring and filtering methods on phylogenetic and evolutionary
analyses has not been well studied. Even for a single filtering
program, Gblocks, results have been contradictory: one
simulation-based study found that it improved the phylogenetic signal
\citep{Talavera2007Improvement} while an empirical study across a wide
range of taxa found that Gblocks-filtered alignments produced worse
phylogenetic trees than unfiltered alignments
\citep{Dessimoz2010Phylogenetic}. A recent study using a variety of
filters suggested that the benefit of alignment filtering (in terms of
improved accuracy) outweighs the cost (in terms of reduced power) when
applied to detecting positive selection \citep{Privman2011Improving},
but this analysis was limited to a small range of possible
evolutionary scenarios as discussed below.  With the application of
published filtering methods to alignments before testing for positive
selection becoming standard practice
\citep{Studer2008Pervasive,Aguileta2009Rapidly}, continued
investigation of potential benefits of alignment filtering to the
detection of positive selection seems well-warranted.

This paper aims to use a simulation framework to incorporate alignment
error and alignment filtering into estimates of the error rate and
power of \sw evolutionary inference of positive selection. Our
approach builds on those of Anisimova et al.
\citeyearpar{Anisimova2002Accuracy}, Fletcher and Yang
\citeyearpar{Fletcher2010Effect}, and Privman et
al. \citeyearpar{Privman2011Improving}, using simulated protein
alignments including insertions and deletions to evaluate methods for
detecting \sw positive selection. Furthermore, we incorporate a
diverse sample of aligners and alignment filters into an experimental
design that differs from previous ones in a number of important ways.

We focus on sitewise detection of positive selection occurring
throughout a phylogeny and evaluate the impact of a number of
alignment filtering methods on the \sw analysis. Thus, the biological
hypothesis being investigated is different from that studied by
Fletcher and Yang \citeyearpar{Fletcher2010Effect}, who focused on
genewise selection acting at specific branches.  Privman et
al. \citeyearpar{Privman2011Improving} recently published a related
paper considering the evolutionary chracteristics of three HIV-1
genes. They concluded that alignment filtering improves the
performance of positive selection inference by reducing false positive
results. While this may be true for these three genes (and perhaps for
HIV-1 in general), HIV-1 is known to evolve with widespread positive
selection in the human host \citep{Yang2003Widespread}. Results valid
for these genes may not be widely applicable to large-scale vertebrate
and mammalian comparative datasets, which exhibit less adaptive
evolution \citep{Kosiol2008Patterns} and which comprise a larger
diversity of protein structures and a wider range of species
divergence levels.

We hypothesize that the divergence level and indel rate, two important
evolutionary factors which are highly variable within and between
different genomes, may strongly affect the performance of methods for
alignment and detection of selection. Accordingly, our simulations
encompass a wide range of biologically plausible indel rates and
divergence levels while fixing other parameters at values typical to
those encountered in the \sw analysis of vertebrate gene families.

@@

\begin{table}[p]
\caption{Genome-wide Divergence Estimates for Commonly Analyzed Eukaryotes}
\begin{center}
\begin{threeparttable}

\defcitealias{Hillier2004Sequence}{}
\defcitealias{Siepel2005Evolutionarily}{}

\footnotesize \tabcolsep 3pt
\begin{tabular}{
l r r l
}
\toprule
 Species
 & Pairwise $dS$
 & Root-to-tip $dS$
 & Reference
\\
\midrule
   Human-Chimp & 0.01 & (0.005) & \citealp{Nei2010Neutral}
\\ Human-Mouse & 0.43 & (0.215) & \citealp{Nei2010Neutral}
\\ Human-Mouse & 0.5 - 0.8 & (0.25 - 0.4) & \citealp{Ogurtsov2004Indel}
\\ Human-Chicken & 0.9 & (0.45) & \citealp{Nei2010Neutral}
\\ Human-Chicken & 1.66 & (0.83) & Hillier et al., 2004 \citetalias{Hillier2004Sequence}
\\ Human-Zebrafish & 1.38 & (0.69) & \citealp{Nei2010Neutral}
\\ Vertebrates & -- & 0.75 & Siepel et al., 2005 \citetalias{Siepel2005Evolutionarily} 
\\ Drosophila & -- & 1.0 & Siepel et al., 2005
\\ Yeasts & -- & 1.25 & Siepel et al., 2005
\\
\bottomrule
\label{table_2}
\end{tabular}
\begin{tablenotes}

\item [] N{\sc ote}.---{\hskip -.1pt}The root-to-tip $dS$ is
equivalent to the MPL (mean path length) used in our simulations. For
two-species comparisons where the pairwise $dS$ was given, the
root-to-tip $dS$ was calculated as half of the pairwise $dS$ and is
included in parentheses.

\end{tablenotes}
\end{threeparttable}
\end{center}
\end{table}

@@

An overview of the simulation parameters used in this study can be
found in Table \ref{table_1}. Three rooted trees were used to guide
the simulation of protein-coding DNA alignments: the artificial
6-taxon tree used by Anisimova et al.
\citeyearpar{Anisimova2001Accuracy} and Massingham and Goldman
\citeyearpar{Massingham2005Detecting} rooted at its midpoint, the
17-taxon vertebrate $\beta$-globin tree from Yang et
al. \citeyearpar{Yang2000CodonSubstitution} and the 44-taxon
vertebrate tree used by the ENCODE project
\citep{2007Identification,Nikolaev2007Early}. Trees, shown with their
original branch lengths in Figure \ref{fig_1}, were scaled to
comparable divergence levels by normalizing their mean path length
(MPL), defined as the root-to-tip branch length averaged across all
lineages in the tree. We simulated alignments with MPL divergence
between 0.05 and 2.0 \syn substitutions per \syn site,
spanning the range of evolutionary divergences observed in several
clades of organisms with fully-sequenced genomes (Table
\ref{table_2}).
\begin{figure}[t]
\begin{center}
\hidegraphics{scale=0.8}{fig1.pdf}
\end{center}
\caption{Phylogenetic trees used for simulation and analysis. The
  original scale for each tree is indicated by a scale bar, but trees
  were scaled to equal mean path length (MPL) divergence levels for
  simulation. (A) A 6-taxon artificial tree used in previous
  simulations
  \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (B) A tree
  estimated from $\beta$-globin genes of 17 vertebrates and used in
  previous empirical analyses and simulation studies
  \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (C) The
  44-species tree used by the ENCODE project
  \citep{2007Identification,Nikolaev2007Early}. The nodes indicated by
  arrows were used as the reference species when comparing the true
  and inferred alignment (see Methods).}
\label{fig_1}
\end{figure}

@@

The distribution of \sw selective pressures (embodied by the parameter
$\omg$, the ratio of the rate of \nsyn substitution to the
rate of \syn substitution) was modeled with a log-normal
distribution derived from a maximum-likelihood fit to a large dataset
of \sw selective pressures estimated from mammalian gene trees
(Lindblad-Toh et al. \citeyear{LindbladToh2011High}; log-normal parameters shown in Table
\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
sites having $\omg>1$, is consistent with the structure-based
expectation of many protein sites under purifying selection and few
under neutral selection or positive selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. INDELible's general
discrete model of \sw $\omg$ variation was used to approximate the
log-normal distribution by splitting the probability density into 50
equally-spaced bins between $\omg$ values of 0 and 3, with the highest
bin containing the probability density for all values $\omg>3$.

Branch lengths for each of the simulation trees were scaled before
simulation to correct for the difference between our definition of
branch lengths as the number of \syn substitutions per
\syn site ($dS$) and INDELible's interpretation of branch length
as the average number of substitutions per codon ($t$)
\citep{Fletcher2010Effect}. They are related approximately by
$t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and $S$ are the
proportion of \nsyn and \syn sites and $\bar{\omg}$ is
the mean $\omg$ across all sites. $S$ is approximately 0.3 when
$\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio
for our chosen distribution is 0.277, yielding a $dS$-to-$t$
conversion factor of 1.48 for all simulations performed.

@@

Alignments were inferred using six alignment algorithms chosen for
their widespread use or demonstrated accuracy: ClustalW v1.82
\citep{Thompson1994ClustalW}, MAFFT \citep{Katoh2005MAFFT}, \tcr{ProbCons
\citep{Do2005a}, T-Coffee \citep{Notredame2000TCoffee}} and two variants of
PRANK \citep{Loytynoja2008PhylogenyAware} based on an amino acid model
(subsequently referred to as \pranka{}) or an empirical codon model
(subsequently referred to as \prankc{}). Unaligned amino acid
sequences were given as input to all alignment programs (except
\prankc{}, which was provided the unaligned DNA sequences) and all
software was run using default parameters with the true phylogenetic
tree given as input where possible.

@@

GUIDANCE and T-Coffee filters produce scores for each residue,
allowing individual residues to be masked instead of entire
columns. Privman et al. \citeyearpar{Privman2011Improving} found a
residue-based filter to be more effective than its column-based
equivalent, and we opted to filter residues instead of alignment
columns where possible. GUIDANCE generates many replicate alignments,
each using a slightly perturbed guide tree, with either MAFFT or
\pranka as the bootstrap aligner. The program then assigns to each
residue from the input alignment a score from 0 to 1 based on how
consistently it was placed in the replicate alignments. In order to
maximize the similarity between the input aligner and the bootstrap
aligner, we ran GUIDANCE with 100 MAFFT replicates when filtering
ClustalW alignments and with 30 \pranka replicates when filtering
\prankc alignments. T-Coffee calculates the residue-wise consistency
between an input multiple alignment and independently calculated
pairwise alignments \citep{Notredame2003Using}, rounding and
normalizing residue scores into integers between 0 and 9. T-Coffee was
run using its default settings and the {\em evaluate\_mode
  --output=score\_ascii} command-line parameters to output alignment
scores.

@@

\begin{figure}[t]
\begin{center}
\hidegraphics{scale=1.25}{fig2.pdf}
\end{center}
\caption{Alignments were simulated without indels for three tree
  shapes and analyzed with SLR, \meight, or \mtwo. Fifty replicate
  alignments were simulated for each data point. The performance of
  each analysis method, as measured by four summary statistics, is
  plotted as a function of the mean path length (MPL) divergence. From
  top to bottom: true positive rate (TPR) at the recommended cutoff
  threshold (0.95 for PAML and 3.84 for SLR); false positive rate
  (FPR) at the recommended cutoff threshold; TPR at a 1\% FPR
  threshold; Pearson's correlation coefficient between the true and
  inferred sitewise $\omg$.}
\label{fig_2}
\end{figure}

\section*{Results and Discussion}
\subsection*{The Performance of Three Methods for Detecting \Sw Positive Selection}
We first evaluated the ability of three \sw methods, \mtwo, \meight
and SLR, to accurately estimate \sw $\omg$ values and to detect
positive selection under a range of tree lengths in the absence of
alignment error. Figure \ref{fig_2} shows the TPR, FPR, \tpr{} and \sw
$\omg$ correlation over a range of mean path lengths
\tcr{(MPL, defined as the mean root-to-tip branch
  length across all lineages)} for each of the three simulation trees.

The detection power and $\omg$ correlation were weakest at low
divergence levels for all methods and all trees due to the low amount
of evolutionary information, as observed in previous simulations
\citep{Anisimova2002Accuracy}. We found a positive correlation between
tree size and detection power, with the highest performance in the
44-taxon tree. Power generally increased monotonically with
divergence, except for the 6-taxon tree which saw its maximum
performance at moderate divergence levels (MPL 0.5--1.0) and began
decreasing at higher values. The downward trend in the 6-taxon tree
was likely due to the impact of saturation of \syn sites in the
very long branches present in such a sparse tree at high divergence
levels. With lower average branch lengths at equivalent MPLs, the two
larger trees showed no signs of decreased performance even at a MPL of
2 substitutions per site, which is greater than any of the divergence
levels found in groups of commonly analyzed vertebrate, insect and
fungal species (Table \ref{table_2}).

@@

In this paper, we investigated the performance of sitewise detection
of positive selection under a range of tree sizes, indel rates, and
divergence levels, using simulation parameters designed to approximate
the analysis of typical mammalian protein-coding genes. We
evaluated the ability of six alignment methods and three alignment
filtering methods to produce alignments for detecting positively
selected sites, using the FPR, TPR, and error-controlled \tpr of the
sitewise detection of positive selection as our main measures of
performance.
